State of the Practice for Lattice Boltzmann Method Software

### 2022-01-04 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex

\begin{table}
\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\hline
		Name & Released & Updated\\
		\hline
		DL\_MESO (LBE) & unclear & 2020 Mar\\
		ESPResSo & 2010 Nov & 2020 Jun\\
		ESPResSo++ & 2011 Feb & 2020 Apr\\
		lbmpy& unclear  & 2020 Jun  \\
		lettuce & 2019 May & 2020 Jul\\
		Ludwig& 2018 Aug & 2020 Jul\\
		LUMA& 2016 Nov   & 2020 Feb\\
		MechSys & 2008 Jun    & 2021 Oct\\
		Musubi & 2013 Sep & 2019 Aug\\
		OpenLB & 2007 Jul & 2019 Oct\\
		Palabos & unclear & 2020 Jul\\
		pyLBM & 2015 Jun&   2020 Jun\\
		Sailfish & 2012 Nov & 2019 Jun\\
		TCLB & 2013 Jun  & 2020 Apr\\
		waLBerla & 2008 Aug & 2020 Jul\\
		\hline
	\end{tabular}
	\caption{Alive Software Packages} \label{alivepackages}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\hline
		Name & Released & Updated\\
		\hline
		HemeLB & 2007 Jun & 2018 Aug\\
		laboetie & 2014 Nov & 2018 Aug\\		
		LatBo.jl & 2014 Aug & 2017 Feb\\
		LB2D-Prime & 2005 & 2012 Apr\\
		LB3D & unclear & 2012 Mar\\
		LB3D-Prime & 2005 & 2011 Oct\\
		LIMBES & 2010 Nov & 2014 Dec\\
		MP-LABS & 2008 Jun & 2014 Oct\\
		SunlightLB & 2005 Sep & 2012 Nov\\
		\hline
	\end{tabular}
	\caption{Dead Software Packages} \label{deadpackages}
\end{center}
\end{table}

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{3cm}llllllll}
			\hline
			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
			\hline
			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			\hline
		\end{tabular}
		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3),
			Pll for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or No), Rflx for
			Reflexive Boundary Condition (Yes or No), MFl for Multi-fluid (Yes or
			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Windows (W), macOS (M), Linux (L)))} \label{tbl_features}
	\end{center}
\end{table}

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{3.5cm}p{3.5cm}p{3.5cm}p{2.5cm} }
			\hline
			Name & $\%$ Issues Closed & $\%$ Code Comments & Status\\
			\hline
			DL\_MESO (LBE) & Not Git & 8.06& Alive\\
			ESPResSo & 89.26 & 21.78 & Alive\\
			ESPResSo++ & 66.28 & 17.10 & Alive\\
			HemeLB & No Issues & 16.68 & Dead\\
			laboetie & 18.75 & 2.47 & Dead\\		
			LatBo.jl & 93.33 & 0.40 & Dead\\
			LB2D-Prime & Not Git & 13.61 & Dead\\
			LB3D & Not Git & 13.76 & Dead\\
			LB3D-Prime & Not Git & 14.34 & Dead\\
			lbmpy& 58.33  & 2.03 & Alive\\
			lettuce & 33.33 & 8.19 & Alive\\
			LIMBES & Not Git & 17.39 & Dead\\
			Ludwig& 60.00 & 20.70 & Alive\\
			LUMA& 85.71   & 0.20 & Alive\\
			MechSys & Not Git & 15.11 & Alive\\
			MP-LABS & 100.00 & 26.67 & Dead\\
			Musubi & Not Git & 24.19 & Alive\\
			OpenLB & Not Git & 22.43 & Alive\\
			Palabos & 89.47 & 17.76 & Alive\\
			pyLBM & 66.67& 16.12 & Alive\\
			Sailfish & 22.22 & 9.26 & Alive\\
			SunlightLB & Not Git & 17.67 & Dead\\
			TCLB & 60.32 & 6.02 & Alive\\
			waLBerla & 72.90 & 22.62 & Alive\\
			\hline
		\end{tabular}
		\caption{Git Repository Data} \label{gitrepodata}
	\end{center}
	\end{table}

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{3.5cm}p{2cm}p{2.5cm}p{2cm}p{2.5cm} }
			\hline
			Name & Text Files & Binary Files & LOC & Avg. LOC / Text File\\
			\hline
			DL\_MESO (LBE) & 310 & 51 & 170223 & 549\\
			ESPResSo & 1390& 83 & 195083 & 140\\
			ESPResSo++ & 1406& 30 & 165194 & 118\\
			HemeLB & 1102& 44 & 123806 & 112\\
			laboetie & 133& 1 & 48403 & 364\\		
			LatBo.jl & 41& 0 & 42172 & 1029\\
			LB2D-Prime & 82& 19 & 54755 & 668\\
			LB3D & 99 & 76 & 39766 & 402\\
			LB3D-Prime & 23& 6 & 12944 & 563\\
			lbmpy& 250 & 29 & 61632 & 247\\
			lettuce & 73 & 1 & 7660 & 105\\
			LIMBES & 26 & 1 & 4872 & 187\\
			Ludwig & 954 & 32 & 162518 & 170\\
			LUMA & 314 & 19 & 4399723 & 14011\\
			MechSys & 333 & 3 & 95707 & 287\\
			MP-LABS & 307 & 3 & 43124 & 140\\
			Musubi & 1347 & 1839 & 281879 & 209\\
			OpenLB & 1438 & 7 & 218406 & 152\\
			Palabos & 1974 & 71 & 563841 & 286\\
			pyLBM & 272 & 108 & 37234 & 137\\
			Sailfish & 632 & 11 & 69398 & 110\\
			SunlightLB & 36 & 1 & 7646 & 212\\
			TCLB & 594 & 7 & 49156 & 83\\
			waLBerla & 2643 & 69 & 873988 & 331\\
			\hline
		\end{tabular}
		\caption{Module Data} \label{moduledata}
	\end{center}
\end{table}

@@

, without a need for a rigorous software development model. The developer of
ESPResSo noted that while no formal model is used, their development model is
something similar to a combination of agile and waterfall development models.

@@

\subsection{Artifacts Present} \label{artifacts}

This subsection answers the research question: What artifacts are present in
current software packages?

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{8 cm} }
			\hline
			Common\\
			\hline
			Authors / Developers List\\
			Bug Tracker\\
			Dependency Notes\\
			Installation Guide / Instructions\\
			License\\
			List of Related Articles / Publications\\
			Makefile / Build File\\
			README File\\
			Requirements Specification / Theory Notes\\
			Tutorial\\
			\hline
			Less Common\\
			\hline
			Change Log / Release Notes\\
			Design Documentation\\
			Functional Specification / Notes\\
			Performance Information / Notes\\
			Test Plan / Report / Script / Data / Cases\\
			User Manual/Guide\\
			Version Control\\
			\hline
			Rare\\
			\hline
			API Documentation\\
			Developer / Contributor Manual / Guide\\
			FAQ / Forum\\
			Verification and Validation Plan / Notes\\
			Video Guide (including YouTube)\\
			\hline
		\end{tabular}
		\caption{Artifacts Present} \label{artifactspresent}
	\end{center}
\end{table}

\subsubsection{Common Artifacts}

@@

The common artifacts contribute to the quality of the software in the following
ways. A list of authors and developers helps potential users and contributors
contact project members to answer questions affecting many software qualities. A
bug tracker helps with organizing improvements to the software. Dependency notes
help users install the software, as does an installation guide. Makefiles or
other automated build files decrease human error in the installation process.
Licenses promote usage of the software. Requirements specifications, linked
theory notes, and related articles and publications help make the software more
understandable, decrease ambiguity, and help with verifying its correctness.
README files also help with understandability. The tutorials help users become
familiar with using the software.

\subsubsection{Less Common Artifacts} \label{lesscommon}

The top four AHP ranked packages have most of the less common artifacts. At the
time of data collection, only one (Palabos) of the four packages did not have a
user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. This broken link was later fixed,
but this is not reflected in our data because it was not present at the time of
data collection. Despite the broken link, Palabos does have a detailed and
informative website. Another one of the top four packages, Ludwig, does not
appear to have publicly visible design documentation. A third package from the list, OpenLB, does not appear to use a version control system. It is possible
that such a system is used by the package since its website notes package
version numbers, but the artifacts do not explicitly state the use of such a
system.

The less common artifacts also contribute to software quality. A change log or
release notes improve traceability of the software. Design documentation helps
with maintaining, modifying, and reusing the software. Version control also
helps to improve these qualities, as well as with traceability. Functional
specifications and notes clarify the software, contributing to usability and
understandability. A user manual also helps with those qualities and with
installability, and correctness and verifiability. Performance notes suggest
that performance was considered when developing the software. Test plans,
scripts, and cases, help verify correctness.

\subsubsection{Rare Artifacts}

It is not common for the top four AHP ranked packages to have many of the rare
artifacts. None of the top four packages have any explicit API documentation.
Three of these packages (ESPREsSo, Ludwig, Palabos) have information on
contributing to the project. Two of them (OpenLB, Palabos) have a FAQ section or
forum. One (OpenLB) has verification and validation notes, and a video guide of
the software.

Although the artifacts were rarely found in our set of LBM software, they also
contribute to software quality. API documentation helps with reusing the
software. Developer and contributor manuals and guides help with
maintainability, visibility and transparency. FAQs and forums improve usability
of software. Verification and validation notes can be used to help check if a
system meets specifications. Video guides can contribute to many software
qualities, depending on the content of the video.

\subsection{Tools Used} \label{tools}

This subsection answers the research question: What tools (development,
dependencies, project management) are used by current software packages?

Software tools are used to support the development, verification, maintenance,
and evolution of software, software processes, and artifacts
\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages.
The tools noted here are subdivided into development tools, dependencies, and
project management tools.

\subsubsection{Development Tools}

Development tools support the development of end products, but do not become
part of them, unlike dependencies that remain in the application once it is
released \citep{ghezzi1991fundamentals}. The following type of development tools
were explicitly noted in the artifacts or web-pages of the 24 LBM packages that
were assessed. It is likely that other tools, such as debuggers, were used but
are not specified in our sources.

	\begin{multicols}{2}	
		\begin{itemize}

			\item Continuous Integration Tools
			\item Code Editors
			\item Development Environment
			\item Runtime Environments
			\item Compilers
			\item Unit Testing Tools
			\item Correctness Verification Tools

		\end{itemize}
	\end{multicols}

The above tools can verify the correctness of software during its development.
Only three (ESPResSo, Ludwig, Musubi) of the software packages that were assessed mentioned using continuous integration tools, like Travis CI. Code editors and compilers were explicitly noted to have been used by several packages, and were likely used by all of them. One of the packages (Ludwig) explicitly noted the use of proprietary unit testing code written in C. Likewise, the use of proprietary code for verifying the correctness of output was noted by one (pyLBM) of the developers. It is likely that similar tools were used when developing other software packages.

\subsubsection{Dependencies}

The following types of dependencies were explicitly noted in the artifacts or
web-pages of the 24 LBM packages that were assessed. It is possible that other
types of dependencies are part of these software packages, but are not clearly
specified in their artifacts or web sites and because of that they are not
listed here.

	\begin{multicols}{2}	
		\begin{itemize}

			\item Build Automation Tools
			\item Technical Libraries
			\item Domain Specific Libraries

		\end{itemize}
	\end{multicols}

Most of the software packages use some sort of build automation tools, most
commonly Make. They also all use various technical and domain specific
libraries. Technical libraries include visualization (e.g. Matplotlib, ParaView,
Pygame, VTK), data analysis (e.g. Anaconda, Torch), and message passing
libraries (e.g. MPICH, Open MPI, PyZMQ). Domain specific libraries are
scientific computing libraries (e.g. SciPy). Libraries that are not explicitly
stated in artifacts, or were not noted during our observations, may fall outside
of these categories.

\subsubsection{Project Management Tools}

Many of the software packages that were assessed were developed by teams of two
or more people. Their work needed to be coordinated and managed. The following
types of project management tools were explicitly noted in the artifacts,
web-pages, or interviews with the developers of the 24 LBM packages that were
assessed. As with development tools and dependencies, it is possible that other
types of project management tools were used to coordinate and manage the
projects but are not specified and because of that they are not listed here.

	\begin{multicols}{2}	
		\begin{itemize}

			\item Collaboration Tools
			\item Email
			\item Change Tracking Tools
			\item Version Control Tools
			\item Document Generation Tools

		\end{itemize}
	\end{multicols}

Collaboration tools are noted as being used when developing the software
projects. Most often email and video conferencing is used. Project management
software was not explicitly mentioned, but it is possible that some of the
projects use such software. Many of the projects are located on GitHub, and its
developers use the platform to help manage their projects, especially bug
related issues. Most of the projects appear to use change tracking and version
control tools. They often use GitHub or GitLab for this. One package
(SunlightLB) uses CVS. Document generation tools are mentioned in the artifacts
of 13 of the projects. The tools Sphinx and Doxygen are explicitly used in this
capacity.

\subsection{Principles, Processes, and Methodologies} \label{prinprocmeth}

This subsection answers the research question: What principles, processes, and
methodologies are used in the development of current software packages?

The points and conclusions in this subsection come from developer interviews and
reviews of software package artifacts.

@@

Processes, like methods, are ways of doing things, especially in an orderly way;
while methodologies are defined as systems of methods
\citep{ghezzi1991fundamentals}. It is not explicitly indicated in the artifacts
of most of the packages that development involved following any specific model,
like a waterfall or agile development model. One developer (ESPResSo) noted that
while no formal model is used, their development model is something similar to a
combination of agile and waterfall development models. The developer teams of
the LBM packages are fairly small, so it is feasible for them to be organized
without the need for such models.

@@

\subsection{Quality Recommendations} \label{qualityrecommentations}

@@

\subsection{Designation Comparison} \label{comparison}

@@

\begin{table}
\begin{center}
		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
			\hline
			Name & Our Ranking & Repository Stars & Repository Star Rank & Repository Watches \\
			\hline
			ESPResSo &1& 145 &2& 19\\
			Ludwig&2& 27 &8& 6\\
			Palabos &3& 34 &6& GitLab\\
			OpenLB &4& No Git &No Git & No Git\\
			LUMA&5& 33 &7& 12\\
			pyLBM &6& 95&3& 10\\
			DL\_MESO (LBE) & 7 & No Git &No Git & No Git \\
			Musubi & 8 & No Git &No Git & No Git \\
			Sailfish &9& 186 &1& 41\\
			waLBerla & 10 & 20 &9& GitLab\\
			laboetie &11& 4 &13& 5\\
			TCLB &12& 95 &3& 16\\
			MechSys &13& No Git &No Git & No Git\\
			lettuce &14& 48 &4& 5\\
			ESPResSo++ &15& 35 &5& 12\\
			MP-LABS &16& 12 &11& 2\\			
			SunlightLB & 17& No Git &No Git & No Git\\
			LB3D &18& No Git &No Git & No Git\\			
			LIMBES &19& No Git &No Git & No Git\\
			LB2D-Prime &20& No Git &No Git & No Git\\		
			HemeLB &21& 12 &11& 12\\
			lbmpy&22&  11 &12&  2\\	
			LB3D-Prime &23& No Git &No Git & No Git\\	
			LatBo.jl &24& 17 &10& 8\\			
			\hline
		\end{tabular}
		\caption{Repository Ranking Metrics} \label{repometrics}
\end{center}
\end{table}

@@

Regarding data collection, a risk to the validity of this assessment is missing
or incorrect data. Some software package data may not have been measured due to
technology issues like broken links. This issue arose with the measurement of
Palabos, which had a broken link to its user manual, as noted in Section
~\ref{lesscommon}.

@@

>>>>

\begin{table}
\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\toprule
		Name & Released & Updated\\
		\midrule
		DL\_MESO (LBE) & unclear & 2020 Mar\\
		ESPResSo & 2010 Nov & 2020 Jun\\
		ESPResSo++ & 2011 Feb & 2020 Apr\\
		lbmpy& unclear  & 2020 Jun  \\
		lettuce & 2019 May & 2020 Jul\\
		Ludwig& 2018 Aug & 2020 Jul\\
		LUMA& 2016 Nov   & 2020 Feb\\
		MechSys & 2008 Jun    & 2021 Oct\\
		Musubi & 2013 Sep & 2019 Aug\\
		OpenLB & 2007 Jul & 2019 Oct\\
		Palabos & unclear & 2020 Jul\\
		pyLBM & 2015 Jun&   2020 Jun\\
		Sailfish & 2012 Nov & 2019 Jun\\
		TCLB & 2013 Jun  & 2020 Apr\\
		waLBerla & 2008 Aug & 2020 Jul\\
		\bottomrule
	\end{tabular}
	\caption{Alive Software Packages} \label{alivepackages}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
	\begin{tabular}{ p{4cm}p{3cm}p{3cm} }
		\toprule
		Name & Released & Updated\\
		\midrule
		HemeLB & 2007 Jun & 2018 Aug\\
		laboetie & 2014 Nov & 2018 Aug\\		
		LatBo.jl & 2014 Aug & 2017 Feb\\
		LB2D-Prime & 2005 & 2012 Apr\\
		LB3D & unclear & 2012 Mar\\
		LB3D-Prime & 2005 & 2011 Oct\\
		LIMBES & 2010 Nov & 2014 Dec\\
		MP-LABS & 2008 Jun & 2014 Oct\\
		SunlightLB & 2005 Sep & 2012 Nov\\
		\bottomrule
	\end{tabular}
	\caption{Dead Software Packages} \label{deadpackages}
\end{center}
\end{table}

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{3cm}llllllll}
			\toprule
			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
			\midrule
			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
			Musubi & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			\bottomrule
		\end{tabular}
		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3),
			Pll for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or No), Rflx for
			Reflexive Boundary Condition (Yes or No), MFl for Multi-fluid (Yes or
			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Windows (W), macOS (M), Linux (L)))} \label{tbl_features}
	\end{center}
\end{table}

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{3.5cm}p{3.5cm}p{3.5cm}p{2.5cm} }
			\toprule
			Name & $\%$ Issues Closed & $\%$ Code Comments & Status\\
			\midrule
			DL\_MESO (LBE) & Not Git & 8.06& Alive\\
			ESPResSo & 89.26 & 21.78 & Alive\\
			ESPResSo++ & 66.28 & 17.10 & Alive\\
			HemeLB & No Issues & 16.68 & Dead\\
			laboetie & 18.75 & 2.47 & Dead\\		
			LatBo.jl & 93.33 & 0.40 & Dead\\
			LB2D-Prime & Not Git & 13.61 & Dead\\
			LB3D & Not Git & 13.76 & Dead\\
			LB3D-Prime & Not Git & 14.34 & Dead\\
			lbmpy& 58.33  & 2.03 & Alive\\
			lettuce & 33.33 & 8.19 & Alive\\
			LIMBES & Not Git & 17.39 & Dead\\
			Ludwig& 60.00 & 20.70 & Alive\\
			LUMA& 85.71   & 0.20 & Alive\\
			MechSys & Not Git & 15.11 & Alive\\
			MP-LABS & 100.00 & 26.67 & Dead\\
			Musubi & Not Git & 24.19 & Alive\\
			OpenLB & Not Git & 22.43 & Alive\\
			Palabos & 89.47 & 17.76 & Alive\\
			pyLBM & 66.67& 16.12 & Alive\\
			Sailfish & 22.22 & 9.26 & Alive\\
			SunlightLB & Not Git & 17.67 & Dead\\
			TCLB & 60.32 & 6.02 & Alive\\
			waLBerla & 72.90 & 22.62 & Alive\\
			\bottomrule
		\end{tabular}
		\caption{Git Repository Data} \label{gitrepodata}
	\end{center}
	\end{table}

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{3.5cm}p{2cm}p{2.5cm}p{2cm}p{2.5cm} }
			\toprule
			Name & Text Files & Binary Files & LOC & Avg. LOC / Text File\\
			\midrule
			DL\_MESO (LBE) & 310 & 51 & 170223 & 549\\
			ESPResSo & 1390& 83 & 195083 & 140\\
			ESPResSo++ & 1406& 30 & 165194 & 118\\
			HemeLB & 1102& 44 & 123806 & 112\\
			laboetie & 133& 1 & 48403 & 364\\		
			LatBo.jl & 41& 0 & 42172 & 1029\\
			LB2D-Prime & 82& 19 & 54755 & 668\\
			LB3D & 99 & 76 & 39766 & 402\\
			LB3D-Prime & 23& 6 & 12944 & 563\\
			lbmpy& 250 & 29 & 61632 & 247\\
			lettuce & 73 & 1 & 7660 & 105\\
			LIMBES & 26 & 1 & 4872 & 187\\
			Ludwig & 954 & 32 & 162518 & 170\\
			LUMA & 314 & 19 & 4399723 & 14011\\
			MechSys & 333 & 3 & 95707 & 287\\
			MP-LABS & 307 & 3 & 43124 & 140\\
			Musubi & 1347 & 1839 & 281879 & 209\\
			OpenLB & 1438 & 7 & 218406 & 152\\
			Palabos & 1974 & 71 & 563841 & 286\\
			pyLBM & 272 & 108 & 37234 & 137\\
			Sailfish & 632 & 11 & 69398 & 110\\
			SunlightLB & 36 & 1 & 7646 & 212\\
			TCLB & 594 & 7 & 49156 & 83\\
			waLBerla & 2643 & 69 & 873988 & 331\\
			\bottomrule
		\end{tabular}
		\caption{Module Data} \label{moduledata}
	\end{center}
\end{table}

@@

@@

\subsection{What artifacts are present in current software projects?}
\label{artifacts}

@@

\begin{table}
\begin{center}
\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
\toprule
\textbf{Common} & \textbf{Less Common} & \textbf{Rare} \\
\midrule
Authors / Developers List & Change Log / Release Notes & API Documentation\\
Bug Tracker & Design Documentation & Developer / Contributor Manual\\
Dependency Notes & Functional Spec. / Notes & FAQ / Forum\\
Installation Guide / Instructions & Performance Information & Verification and Validation Plan\\
Requirements Spec. / Theory Notes & Test Plan / Script / Cases & Video Guide (including YouTube)\\
List of Related Publications & User Manual/Guide & \\
Makefile / Build File & Version Control & \\
README File & & \\
License & & \\
Tutorial & & \\
\bottomrule
\end{tabular}
\caption{Artifacts Present} \label{artifactspresent}
\end{center}
\end{table}

@@

The top four AHP ranked packages have most of the less common artifacts. At the
time of data collection, only one (Palabos) of the four packages did not have a
user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. This broken link was later fixed,
but this is not reflected in our data because it was not present at the time of
data collection. Despite the broken link, Palabos does have a detailed and
informative website. Another one (Ludwig) of the top four packages does not
appear to have publicly visible design documentation. A third package (OpenLB)
from the list does not appear to use a version control system. It is possible
that such a system is used by the package, since its website notes package
version numbers, but the artifacts do not explicitly state the use of such a
system.

The top four ranked packages do not have many of the rare artifacts. None of the
top four packages have any explicit API documentation. Three of these packages
(ESPREsSo, Ludwig, Palabos) have information on contributing to the project. Two
of them (OpenLB, Palabos) have a FAQ section or forum. One (OpenLB) has
verification and validation notes, and a video guide of the software.

\subsection{What tools (development, dependencies, project management) are used
by current software packages?} \label{tools}

Software tools are used to support the development, verification, maintenance,
and evolution of software, software processes, and artifacts
\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages, as
summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
tools, dependencies, and project management tools.

\begin{table}
	\begin{center}
	\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
	\toprule
	\textbf{Development Tools} & \textbf{Dependencies} & \textbf{Project
	Management Tools} \\
	\midrule
	Continuous Integration & Build Automation Tools & Collaboration Tools\\
	Code Editors & Technical Libraries & Email\\
	Development Environment & Domain Specific Libraries & Change Tracking Tools\\
	Runtime Environments &  & Version Control Tools\\
	Compilers &  & Document Generation Tools\\
	Unit Testing Tools &  & \\
	Correctness Verification Tools &  & \\
	\bottomrule
	\end{tabular}
	\caption{Observed development tools, dependencies and project management tools} \label{tbl_tools}
	\end{center}
\end{table}

Development tools support the development of end products, but do not become
part of them, unlike dependencies that remain in the application once it is
released \citep{ghezzi1991fundamentals}. Although not shown in
Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo, 
Ludwig and Musubi) packages mentioned continuous integration tools, like Travis CI. Code
editors and compilers were explicitly noted to have been used by several
packages, and were likely used by all of them. One of the packages (Ludwig)
explicitly noted the use of proprietary unit testing code written in C.
Likewise, the use of proprietary code for verifying the correctness of output
was noted by one of the developers (pyLBM). Similar tools were likely used when
developing the other software packages.

For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
software packages use some sort of build automation tools, most commonly Make.
They all use various technical and domain specific libraries. Technical
libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
Open MPI, PyZMQ). Domain specific libraries include scientific computing
libraries (e.g.\ SciPy). Some used libraries will be missing, if they are not
explicitly stated in the artifacts we observed.

Many of the software packages that were assessed were developed by teams of two
or more people. Their work needed to be coordinated and managed.
Table~\ref{tbl_tools} shows the types of project management tools that were
explicitly noted in the artifacts, web-pages, or interviews with the developers.
As with development tools and dependencies, it is possible that other types of
project management tools are used, but they were not visible in the artifacts to
which we had access.  Collaboration tools are noted as being used when
developing the software projects. Most often email and video conferencing is
used. Project management software was not explicitly mentioned, but it is
possible that some of the projects use such software. Many of the projects are
located on GitHub, and its developers use the platform to help manage their
projects, especially bug related issues. Most of the projects appear to use
change tracking and version control tools. They often use GitHub or GitLab for
this. One package (SunlightLB) uses CVS. Document generation tools are mentioned
in the artifacts of 12 of the projects. The tools Sphinx and Doxygen are
explicitly used in this capacity. \wss{look at what the other papers say are
commonly used tools and bring that information into this section.}

\subsection{What principles, processes, and methodologies are used?}
\label{prinprocmeth}

@@

Processes, like methods, are ways of doing things, especially in an orderly way;
while methodologies are defined as systems of methods
\citep{ghezzi1991fundamentals}. It is not explicitly indicated in the artifacts
of most of the packages that development involved following any specific model,
like a waterfall or agile development model. One developer (ESPResSo) noted that
while no formal model is used, their development model is something similar to a
combination of agile and waterfall development models. The developer teams of
the LBM packages are fairly small, so it is feasible for them to be organized
without the need for such models.  \wss{some repetition from the previous
section - should try to figure out how to remove the repetition.}

@@

\subsection{What is the current software quality for the projects?}
\label{qualityrecommentations}

@@

\subsection{How does software designated as high quality by this methodology
compare with top-rated software by the community?} \label{comparison}

@@

\begin{table}
\begin{center}
		\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
			\toprule
			Name & Our Ranking & Repository Stars & Repository Star Rank & Repository Watches \\
			\midrule
			ESPResSo &1& 145 &2& 19\\
			Ludwig&2& 27 &8& 6\\
			Palabos &3& 34 &6& GitLab\\
			OpenLB &4& No Git &No Git & No Git\\
			LUMA&5& 33 &7& 12\\
			pyLBM &6& 95&3& 10\\
			DL\_MESO (LBE) & 7 & No Git &No Git & No Git \\
			Musubi & 8 & No Git &No Git & No Git \\
			Sailfish &9& 186 &1& 41\\
			waLBerla & 10 & 20 &9& GitLab\\
			laboetie &11& 4 &13& 5\\
			TCLB &12& 95 &3& 16\\
			MechSys &13& No Git &No Git & No Git\\
			lettuce &14& 48 &4& 5\\
			ESPResSo++ &15& 35 &5& 12\\
			MP-LABS &16& 12 &11& 2\\			
			SunlightLB & 17& No Git &No Git & No Git\\
			LB3D &18& No Git &No Git & No Git\\			
			LIMBES &19& No Git &No Git & No Git\\
			LB2D-Prime &20& No Git &No Git & No Git\\		
			HemeLB &21& 12 &11& 12\\
			lbmpy&22&  11 &12&  2\\	
			LB3D-Prime &23& No Git &No Git & No Git\\	
			LatBo.jl &24& 17 &10& 8\\			
			\bottomrule
		\end{tabular}
		\caption{Repository Ranking Metrics} \label{repometrics}
\end{center}
\end{table}

@@

Regarding data collection, a risk to the validity of this assessment is missing
or incorrect data. Some software package data may not have been measured due to
technology issues like broken links. This issue arose with the measurement of
Palabos, which had a broken link to its user manual, as noted in
Section~\ref{artifacts}.

@@

- change \citet{ghezzi1991fundamentals} to \citet{GhezziEtAl2003} (need to check
each instance, add page numbers).
